Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-08-27 document/ms.tex
updated numbers
expanding discussion of comparison to previous work

\dfmtodo{Include Kipping's comments about eccentricity.}

@@

In \eq{gammaearth}, we define ``Earth analog'' in terms of measurable
quantities with no mention of habitability or composition.
This might seem unsatisfying but the composition of an exoplanet is
notoriously difficult to measure even with large uncertainty and any
definition of habitability is still extremely subjective.
With this in mind, we stick to the observable definition for this \paper.

@@

Our constraint on the rate density of Earth analogs (as defined in
\sect{extrap}) is in tension---even though our result has large fractional
uncertainty---with the result from \citet{petigura}.
This is shown in \fig{real-rate} where we compare the marginalized posterior
probability function for \gammaearth\ to the published value and uncertainty.
Quantitatively, we find that the rate density of Earth analogs is
\begin{eqnarray}\eqlabel{ge-result}
\gammaearth &=& 0.017^{+0.018}_{-0.009}~\densityunit
\end{eqnarray}
where the ``\densityunit'' indicates that this quantity is a rate density, per
natural logarithmic period per natural logarithmic radius.
Converted to these units, \citet{petigura} measured
$0.119_{-0.046}^{+0.035}~\densityunit$ for the same quantity (indicated as the
vertical lines in \fig{real-rate}).
This rate density is \emph{exactly} what Petigura's extrapolation model
predicts but, for comparison, we can also integrate our inferred rate density
over their choice of ``Earth-like'' bin ($200 \le \period/\mathrm{day} < 400$
and $1 \le \radius/\radius_\oplus < 2$) to find a \emph{rate} of
Earth analogs.
The published rate is $0.057_{-0.017}^{+0.022}$ (\citealt{petigura}) and our
posterior constraint is
\begin{eqnarray}
\int_{\period=200\,\mathrm{day}}^{400\,\mathrm{day}}
\int_{\radius=1\,\radius_\oplus}^{2\,\radius_\oplus}
\rate_\ratepars (\ln\period,\,\ln\radius)
\dd[\ln\radius]
\dd[\ln\period]
&=&
0.016_{-0.007}^{+0.011}
\quad.
\end{eqnarray}

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with the previously published result even though we use
\emph{exactly the same dataset}.
This inconsistency is due to the different assumptions made.
In particular, for their analysis, \citet{petigura} asserted that the period
distribution for small planets on long orbits is flat in logarithmic period.
If we enforce the same constraint, we reproduce their result but when we relax
this assumption, allowing the data to constrain the period distribution, we
find evidence for a \emph{decreasing} period distribution as a function of
logarithmic period (see \fig{period}).

@@

Under the assumptions that we have made here, our inference of the occurrence
rate density of exoplanets places a probabilistic constraint on the number of
transiting Earth analogs in the existing \kepler\ dataset.
If we adopt the definition of ``Earth-like'' from \citet[][$200 \le
\period/\mathrm{day} < 400$ and $1 \le \radius/\radius_\oplus < 2$]{petigura},
and integrate the product inferred rate density function and the geometric
transit probability (\eqalt{transitprob}) over this bin, we find that the
expected number of Earth-like exoplanets transiting the stars in the sample of
Sun-like stars chosen by \citet{petigura} is
\begin{eqnarray}\eqlabel{ntransit}
N_{\oplus,\,\mathrm{transiting}} &=& 9.2_{-4.0}^{+5.9}
\end{eqnarray}
where the uncertainties are only on the expectation value and don't include
the Poisson sampling variance.
This is an exciting result because it means that, if we can improve the
sensitivity of exoplanet search pipelines to small planets orbiting on long
periods, then we should find some Earth analogs in the existing data.
Furthermore, because of the treatment of multiple transiting systems in the
catalog, the \True\ expected number of transiting Earth-like exoplanets
orbiting Sun-like stars is probably larger than the values in \eq{ntransit}!

@@

>>>>

Implicit in the expression for the transit probability in \eq{transitprob} is
the assumption that all of the planets are on circular orbits.
Recently, \citet{kipping} demonstrated that when eccentric orbits are
included, our given value is an underestimate by about 10\%.
This effect will propagate directly to our inferred rate densities.
Even though the degeneracy is not exact---due to our choice of priors on the
rate density parameters---it is not a bad approximation to assume that it is
and scale the results down by your preferred factor.
The right thing to do would be to marginalize over this effect directly during
inference but that exercise is beyond the scope of the current \paper.

@@

In \eq{gammaearth}, we use the symbol $\Gamma$ instead of the more commonly
used $\eta$ since we define ``Earth analog'' in terms of measurable quantities
with no mention of habitability or composition.
This might seem unsatisfying but the composition of an exoplanet is
notoriously difficult to measure even with large uncertainty and any
definition of habitability is still extremely subjective.
With this in mind, we stick to the observable definition for this \paper.

@@

Our constraint on the rate density of Earth analogs (as defined in
\sect{extrap}) is in tension---even though our result has large fractional
uncertainty---with the result from \citet{petigura}.
This is shown in \fig{real-rate} where we compare the marginalized posterior
probability function for \gammaearth\ to the published value and uncertainty.
Quantitatively, we find that the rate density of Earth analogs is
\begin{eqnarray}\eqlabel{ge-result}
\gammaearth &=& 0.019^{+0.019}_{-0.010}~\densityunit
\end{eqnarray}
where the ``\densityunit'' indicates that this quantity is a rate density, per
natural logarithmic period per natural logarithmic radius.
Converted to these units, \citet{petigura} measured
$0.119_{-0.046}^{+0.035}~\densityunit$ for the same quantity (indicated as the
vertical lines in \fig{real-rate}).
This rate density is \emph{exactly} what Petigura's extrapolation model
predicts but, for comparison, we can also integrate our inferred rate density
over their choice of ``Earth-like'' bin ($200 \le \period/\mathrm{day} < 400$
and $1 \le \radius/\radius_\oplus < 2$) to find a \emph{rate} of
Earth analogs.
The published rate is $0.057_{-0.017}^{+0.022}$ (\citealt{petigura}) and our
posterior constraint is
\begin{eqnarray}
\int_{\period=200\,\mathrm{day}}^{400\,\mathrm{day}}
\int_{\radius=1\,\radius_\oplus}^{2\,\radius_\oplus}
\rate_\ratepars (\ln\period,\,\ln\radius)
\dd[\ln\radius]
\dd[\ln\period]
&=&
0.019_{-0.008}^{+0.010}
\quad.
\end{eqnarray}

\dfmtodo{Do we want to add a comment about giant planets? Or maybe some sort
of cumulative distribution plot/code snippet?}

\section{Comparison with previous work}

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with previously published results.
In particular, our result is completely inconsistent with the earlier result
based on \emph{exactly the same dataset} (\citealt{petigura}).
This inconsistency is due to the different assumptions made the detailed cause
merits some investigation.
The two key differences between our analysis and previous work are \emph{(a)}
the form of the extrapolation function and \emph{(b)} the presence of
measurement uncertainties on the planet radii.

To make their estimate of \gammaearth, \citet{petigura} asserted a flat
distribution in logarithmic period for small planets.
Our results suggest that the data \emph{do not support} this assumption (see
\fig{period}).
We find that the data require a \emph{decreasing} period distribution in the
relevant range.
A similar result was also found by \citet{dong} and it is apparent in Figure 2
of \citet{petigura}.

To test the significance of the choice of extrapolation function, we relax the
assumption of a uniform period distribution and allow the distribution to be
linear in the same range ($R=1-2\,R_\oplus$ and $P=50-400\mathrm{d}$).
Under this model, the likelihood of the catalog of planets in this range can
be calculated using \eq{poisson-like}.
We apply uniform priors in the physically allowed range of slopes and
intercepts for this distribution and estimate the posterior probability for
the extrapolated rate using MCMC (\citealt{emcee}).
This results give a much more uncertain and substantially lower estimate for
the rate of Earth analogs
\begin{eqnarray}
\Gamma_\oplus &=& 0.072^{+0.088}_{-0.047} \quad.
\end{eqnarray}
With the large error bars, this result is consistent with both results (see
\fig{comparison} where this value is labeled ``linear extrapolation'') but it
does not fully account for the discrepancy.

To examine the effects of measurement uncertainties, we repeat our analysis
with the error bars on the radii artificially set to zero, keeping everything
else the same.
This analysis (labeled ``negligible uncertainties'' in \fig{comparison}) gives
the result
\begin{eqnarray}
\Gamma_\oplus &=& 0.040^{+0.031}_{-0.019} \quad.
\end{eqnarray}
This result is relatively more precise and higher than our final result and
consistent with the value obtained with linear extrapolation.
This confirms the hypothesis that the discrepancy between our result and the
previously published values is the combined result of both of our key
generalizations.

For comparison, we have also included the value of \gammaearth\ implied by
\citet[][their Table 2]{dong}.
This result is based on a power law fit to the period distribution of small
planets ($R=1-2\,R_\oplus$) on long periods ($P=10-250\,\mathrm{d}$) in a
different catalog (\citealt{kepler-catalog}) with a parametric completeness
model.
There are a few factors to consider when comparing to this to our analysis.
Firstly, while \citet{dong} did use a somewhat flexible extrapolation
function than (a power law in log period), it is still very restrictive
considering the large range of periods; a broken power law might have been
more applicable.
Furthermore, their analysis did not incorporate the effects of measurement
uncertainties.
Finally, unlike the \citet{petigura}, the \citet{kepler-catalog} catalog used
by \citet{dong} includes multiple transiting systems.
As mentioned previously, the effect of this selection is hard to determine
without further investigation but it should, intuitively, cause any inference
based on the \citet{petigura} sample an underestimate of the \True\ rate.

@@

Under the assumptions that we have made here, our inference of the occurrence
rate density of exoplanets places a probabilistic constraint on the number of
transiting Earth analogs in the existing \kepler\ dataset.
If we adopt the definition of ``Earth-like'' from \citet[][$200 \le
\period/\mathrm{day} < 400$ and $1 \le \radius/\radius_\oplus < 2$]{petigura},
and integrate the product inferred rate density function and the geometric
transit probability (\eqalt{transitprob}) over this bin, we find that the
expected number of Earth-like exoplanets transiting the stars in the sample of
Sun-like stars chosen by \citet{petigura} is
\begin{eqnarray}\eqlabel{ntransit}
N_{\oplus,\,\mathrm{transiting}} &=& 10.6_{-4.5}^{+5.9}
\end{eqnarray}
where the uncertainties are only on the expectation value and don't include
the Poisson sampling variance.
This is an exciting result because it means that, if we can improve the
sensitivity of exoplanet search pipelines to small planets orbiting on long
periods, then we should find some Earth analogs in the existing data.
Furthermore, because of the treatment of multiple transiting systems in the
catalog, the \True\ expected number of transiting Earth-like exoplanets
orbiting Sun-like stars is probably larger than the values in \eq{ntransit}!

@@

\bibitem[Kipping(2014)]{kipping}
Kipping, D.~M.\ 2014, \arxiv{1408.1393}
